---
title: "Daily Maximum Temperatures in Portland, 1979–2021"
subtitle: "Recreating the NYT Upshot heat-wave chart with real NOAA station data"
format: html
execute:
  echo: false
  warning: false
  message: false
---

In late June 2021, Portland, Seattle, and Vancouver experienced one of the
most extreme heat waves on record for the Pacific Northwest. The chart
below reproduces that moment using 42+ years of real daily maximum
temperature data for Portland International Airport (station
`USW00024229`), pulled live from NOAA's NCEI Access Data Service.

```{r}
#| label: setup
#| include: false

packages <- c("httr", "readr", "dplyr", "ggplot2", "lubridate")
installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

library(httr)
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
```

```{r}
#| label: fetch-data

# Station USW00024229 = Portland International Airport, OR
station_id <- "USW00024229"
start_date <- "1979-01-01"
end_date   <- "2021-06-29"

url <- paste0(
  "https://www.ncei.noaa.gov/access/services/data/v1",
  "?dataset=daily-summaries",
  "&dataTypes=TMAX",
  "&stations=", station_id,
  "&startDate=", start_date,
  "&endDate=", end_date,
  "&units=standard",
  "&format=csv"
)

resp <- GET(url)
stop_for_status(resp)
raw_csv <- content(resp, as = "text", encoding = "UTF-8")
weather <- read_csv(raw_csv, show_col_types = FALSE)
```

```{r}
#| label: clean-data

weather <- weather %>%
  mutate(
    DATE = as_date(DATE),
    fake_date = as_date(paste0("2000-", sprintf("%02d", month(DATE)), "-",
                                sprintf("%02d", day(DATE))))
  ) %>%
  filter(!is.na(TMAX))

last_days <- weather %>%
  filter(DATE >= as_date(end_date) - 2) %>%
  arrange(DATE)

history <- weather %>%
  filter(DATE < as_date(end_date) - 2)
```

```{r}
#| label: fig-portland-temps
#| fig-cap: "Daily maximum temperatures in Portland, 1979-2021, with the late-June 2021 heat wave highlighted in red."
#| fig-width: 10
#| fig-height: 7.5

p <- ggplot() +
  geom_jitter(
    data = history,
    aes(x = fake_date, y = TMAX),
    width = 1.2, height = 0,
    color = "#7a8b99", alpha = 0.10, size = 0.9, shape = 16, stroke = 0
  ) +
  geom_point(
    data = last_days,
    aes(x = fake_date, y = TMAX),
    color = "#b7312c", size = 3.4, shape = 16
  ) +
  annotate(
    "text",
    x = last_days$fake_date[1] - 18,
    y = max(last_days$TMAX) + 5,
    label = "Last 3 days",
    color = "#b7312c", fontface = "bold", hjust = 0, size = 4.6,
    family = "sans"
  ) +
  scale_x_date(
    date_labels = "%b",
    date_breaks = "1 month",
    expand = expansion(mult = c(0.02, 0.04))
  ) +
  scale_y_continuous(
    limits = c(0, 122),
    breaks = seq(20, 120, 20),
    expand = c(0, 0)
  ) +
  coord_cartesian(clip = "off") +
  labs(
    title = paste0("Daily maximum temperatures in Portland, ",
                    year(as_date(start_date)), "-", year(as_date(end_date))),
    subtitle = "120 deg. Fahrenheit",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 14, base_family = "sans") +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey88", linewidth = 0.4),
    axis.text.y = element_text(color = "#6b6f8c", size = 11),
    axis.text.x = element_text(color = "#444444", size = 11, margin = margin(t = 6)),
    axis.ticks = element_blank(),
    axis.line.x = element_line(color = "black", linewidth = 0.6),
    plot.title = element_text(face = "bold", size = 17, color = "#1a1a1a",
                               margin = margin(b = 2)),
    plot.subtitle = element_text(color = "#6b6f8c", size = 11.5,
                                  margin = margin(b = 14)),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(25, 45, 15, 15)
  )

p
```

```{r}
#| label: save-png
#| include: false

ggsave(
  filename = "portland_max_temps_1979_2021.png",
  plot = p,
  width = 10, height = 7.5, dpi = 320,
  bg = "white"
)
```

### About the data

Source: NOAA NCEI GHCN-Daily dataset, station `USW00024229` (Portland
International Airport, OR). Portland's all-time record high — 116°F on
June 28, 2021 — falls within the highlighted three-day window above.

